From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-20 Pi+/paper/reversible1.tex
rev1

\section{A Reversible Programming Language: Set-Theoretic Semantics}~\label{sec:reversibleone}

@@

A natural candidate for a semantic foundation for reversible programming languages is the notion of type
isomorphism. Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667} and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}.

Focusing on finite types, the building blocks of type theory are: the
empty type ($\mathbb{0}$), the unit type ($\mathbb{1}$) containing
just one value $\Acon{tt}$, the sum type ($+$) containing values of
the form $\inlv{v}$ and $\inrv{v}$, and the product ($\times$) type
containing pairs of values $(v_1,v_2)$. One may view each type $A$
as a collection of physical wires that can transmit $|A|$ distinct
values where $|A|$ is a natural number that indicates the size of a
type, computed as: $| \mathbb{0} | = 0$; $| \mathbb{1} | = 1$;
$| A + B | = | A | + |B |$; and
$| A \times B | = | A | * |B |$.  Thus the type
$\mathbb{B} = \mathbb{1} + \mathbb{1}$ corresponds to a wire that can
transmit one of two values, i.e., bits, with the convention that
$\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and
$\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The type
$\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
collection of wires that can transmit three bits. From that
perspective, a type isomorphism between types $A$ and $B$
(such that $|A|=|B|=n$) models a \emph{reversible}
combinational circuit that \emph{permutes} the $n$ different
values. These type isomorphisms are collected in Fig.~\ref{pi-terms}.
\begin{figure}[t]
{\scalebox{\scalef}{$%
%%\noindent\begin{minipage}{.7\linewidth}
\begin{array}{rrcll}
\idc :& A & \iso & A &: \idc \\
\\
\identlp :&  \mathbb{0} + A & \iso & A &: \identrp \\
\swapp :&  A + B & \iso & B + A &: \swapp \\
\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
\identlt :&  \mathbb{1} \times A & \iso & A &: \identrt \\
\swapt :&  A \times B & \iso & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
\absorbr :&~ \mathbb{0} \times A & \iso & \mathbb{0} ~ &: \factorzl \\
\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
  \end{array}$}}

@@

Each line in the top part of the figure introduces a pair of dual constants that witness the type isomorphism in the
middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence
relation by adding the three constructors at the bottom of the figure that witness equivalence and compatible
closure. Although austere, this combinator-based language has the advantage of being amenable to formal analysis for at
least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident connections to type
theory and category theory and hence possesses a rich algebraic structure which we exploit in the remainder of the
paper.

% Specifically, the type isomorphisms of~$\Pi$ are sound and complete for all permutations on finite
% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980}. Algebraically, these types and
% combinators form a \emph{commutative semiring} (up to type isomorphism). Logically, they form a superstructural logic
% capturing space-time trade-offs~\cite{superstructural}. Categorically, they form a \emph{distributive bimonoidal
%   category}~\cite{laplaza72}.

% For the fragment over finite types, the syntax of the language $\Pi$ consists of several sorts:

%  {\scalebox{\scalef}{$%
% \begin{array}{lrcl}
% \textit{Value types} & A,B,C,D &::=& \mathbb{0} \alt \mathbb{1} \alt A+B \alt A\times B \\
% \textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
% \textit{Program types} &&& A \leftrightarrow B \\
% \textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
% \end{array}$}}

\begin{figure}[t]
{\scalebox{\scalef}{%
$\begin{array}{rclcrcl}
\delta(\identlp,~\inrv{v}) & = & v & & \delta(\identrp,~v)        & = & \inrv{v}\\
\delta(\swapp,~\inlv{v})     & = & \inrv{v}\\
\delta(\swapp,~\inrv{v})     & = & \inlv{v}\\
\delta(\assoclp,~\inlv{v})          & = & \inlv{(\inlv{v})} & & \delta(\assocrp,~\inlv{(\inlv{v})}) & = & \inlv{v}         \\
\delta(\assoclp,~\inrv{(\inlv{v})}) & = & \inlv{(\inrv{v})} & & \delta(\assocrp,~\inlv{(\inrv{v})}) & = & \inrv{(\inlv{v})}\\
\delta(\assoclp,~\inrv{(\inrv{v})}) & = & \inrv{v}          & & \delta(\assocrp,~\inrv{v})          & = & \inrv{(\inrv{v})}\\
\delta(\identlt,~(\Acon{tt},v)) & = & v & & \delta(\identrt,~v)             & = & (\Acon{tt},v)\\
\delta(\swapt,~(x,y))         & = & (y,x)    \\
\delta(\assoclt,~(x,(y,z))) & = & ((x,y),z) & & \delta(\assocrt,~((x,y),z)) & = & (x,(y,z))\\
\delta(\dist,~(\inlv{x},z)) & = & \inlv{(x,z)} & & \delta(\factor,~\inlv{(x,z)}) & = & (\inlv{x},z)\\
\delta(\dist,~(\inrv{y},z)) & = & \inrv{(y,z)} & & \delta(\factor,~\inrv{(y,z)}) & = & (\inrv{y},z)
\end{array}$}}
\caption{Semantics of base combinators}\label{fig:delta}
\end{figure}

@@

\noindent A simple denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each
combinator to a permutation between finite sets.

Fig.~\ref{fig:delta} specifies the semantics of the base combinators
using a function $\delta$. It is clear that this function has an inverse $\delta^\dagger$ such that
$\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter in the second figure.

We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:

\[\begin{array}{rcll}
    \sem{A} &=& \mathsf{Fin}~\sizet{A}
\\
    \sem{c : A \isot B} &:& \mathsf{Fin}~\sizet{A} \rightarrow \mathsf{Fin}~\sizet{B} \\
    %% base combinators
    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inlv{v} &=& \sem{c_1}_{A \isot C}~v \\
    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inrv{v} &=& \sizet{A} + \sem{c_2}_{B \isot D}~v \\
    \sem{\idc}~v &=& v \\
    \sem{c_1\fatsemi c_2}~v &=& \sem{c_2} (\sem{c_1}~v) \\
    \sem{c_1\otimes c_2}~(v_1,v_2) &=& (\sem{c_1}~v_1 , \sem{c_2}~v_2)
\end{array}\]

>>>>

\section{A Reversible Programming Language: Set-Theoretic Semantics}
\label{sec:reversibleone}

@@

A natural candidate for a semantic foundation for reversible programming languages is the notion of type
isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}.

Focusing on finite types, the building blocks of type theory are: the empty type ($\zerot$), the unit type
($\onet$) containing just one value $\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$. For this fragment of types, the
syntax of the language $\Pi$ consists of the following sorts:

{\scalebox{\scalef}{$%
\begin{array}{lrcl}
\textit{Value types} & A,B,C,D &::=& \zerot \alt \onet \alt A+B \alt A\times B \\
\textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
\textit{Program types} &&& A \iso B \\
\textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
\end{array}$}}
\begin{figure}[t]
{\scalebox{\scalef}{$%
%%\noindent\begin{minipage}{.7\linewidth}
\begin{array}{rrcll}
\idc :& A & \iso & A &: \idc \\
\\
\identlp :&  \zerot + A & \iso & A &: \identrp \\
\swapp :&  A + B & \iso & B + A &: \swapp \\
\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
\identlt :&  \onet \times A & \iso & A &: \identrt \\
\swapt :&  A \times B & \iso & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
\absorbr :&~ \zerot \times A & \iso & \zerot ~ &: \factorzl \\
\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
  \end{array}$}}

@@

\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$ distinct
values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as: $\sizet{\zerot} = 0$;
$\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and $\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type
$\mathbb{B} = \onet + \onet$ corresponds to a wire that can transmit one of two values, i.e., bits, with the
convention that $\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and $\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The
type $\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires that can transmit three
bits. From that perspective, a type isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a
\emph{reversible} combinational circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a
\emph{commutative semiring} and are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure
introduces a pair of dual constants that witness the type isomorphism in the middle.  These are the \emph{base}
(non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence relation by adding the three
constructors at the bottom of the figure that witness equivalence and compatible closure. Although austere, this
combinator-based language has the advantage of being amenable to formal analysis for at least two reasons: (i) it is
conceptually simple and small, and (ii) it has direct and evident connections to type theory and category theory and
hence possesses a rich algebraic structure which we exploit in the remainder of the paper.

@@

\noindent A denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each combinator to
a bijection between finite sets, that is, a permutation. In this section, we provide a simple semantics expressed in the
metalanguage of set theory. This semantics is then extended to the metalanguage of homotopy type theory in
Sec.~\ref{sec:reversibletwo}.

As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence the denotation of a type $A$ with $n$-elements will be a finite set
$\Fin[n]$ and the denotation of a combinator $c : A \iso B$ such that $\sizet{A} = \sizet{B} = n$ will be a function
from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. In more detail, we enumerate the values $v : A$ where
$\sizet{A} = n$ in a canonical order such that each value $v$ is given an index in $\{ 0,\cdots,n-1 \}$. The enumeration
is defined as follows:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& \emptyset \\
    \mathit{enum}(\onet) &=& \{ \Acon{tt} \} \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda w.\mathit{map}~(\lambda v. (v,w))~\mathit{enum}(A))~\mathit{enum}(B))
\end{array}\]
\noindent For example, the 3 values $\inlv{\Acon{tt}}$, $\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of
type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2 respectively. Given this canonical order, the combinator
$\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ would correspond to the permutation $(1 2 0)$ while the
combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ at the same type would correspond to the
identity permutation.

% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
% the finite set, we can define a denotational semantics which constructs the bijection.
% \note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}
